Imminent range shifts in transboundary fish stocks challenge their management under climate change

### 2021-03-19 Scripts/Manuscript.Rmd
grammar comments

---
title: "Imminent range shifts in transboundary fish stocks challenge their management under climate change"
author: 'Authors: Juliano Palacios-Abrantes^1*^, Thomas L. Frölicher^2,3^, Gabriel Reygondeau^1^, U. Rashid Sumaila^1^, Alessandro Tagliabue^4^, Colette Wabnitz^1,5^, and William W.L. Cheung^1^'
date: 'Affiliations: ^1^Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada. ^2^Climate and Environmental Physics, Physics Institute, University of Bern. ^3^Oeschger Centre for Climate Change Research, University of Bern, Switzerland. ^4^School of Environmental Sciences, University of Liverpool. ^5^Stanford Center for Ocean Solutions, Stanford University, Stanford, United States'
output: bookdown::word_document2
reference_docx:  ../References/sci_template.docx
  # pdf_document: default
bibliography: ../References/ToE_Ref.bib
csl: ../References/science.csl
---

<!-- # Instructions  -->

<!-- Research articles should present a major advance and can be short or long (maximum 15,000 words) format. Submissions must include an abstract (up to 150 words), an introduction, and sections with brief informative subheadings. You may include up to ten figures and/or tables and about 60 references. Supplementary materials should be limited to information that is not essential for the general understanding of the research presented in the main text and can include data sets, figures, tables, videos, or audio files.  You may include up to a total of ten figures and/or tables (combined) throughout the supplemental text.  -->

@@

#### Project's Library
packages <- c(
  "tidyverse" ,
  "png",
  "grid",
  "knitr",
 "kableExtra"
  # "here", # for dbem_import `here()`
  # "data.table", #dbem_import `fread()`
  # "readxl", # for reading excell files
  # "janitor", # for clearing names
  # "geosphere", # estimate distances between points `distm()`
  # "ggrepel",
  # "zoo", # for average mean
  # "parallel", # for mclapply,
  # "sf", # for mapping
  # "st", # for mapping
  # "rgdal", #Spatial analysis
  # "tools", #Spatial analysis 
  # "zeallot", # for Juanito's map
  # "gmt", # for estimating distances between points
  # "viridis"
)

@@

corresponding author: [j.palacios\@oceans.ubc.ca](mailto:j.palacios@oceans.ubc.ca){.email}

**Abstract**

<!-- An opening sentence that states the question/problem addressed by the research AND Enough background content to give context to the study AND A brief statement of primary results AND A short concluding sentence. -->

Climate change is shifting the distribution of transboundary fish stocks between neighbouring countries. The time-frame of such shifts determine the extent of climate change challenges to international fisheries governance. Using global climate-fisheries model simulations, we show that, within the next 15 years, 68% of the world’s transboundary fish stocks are projected to shift their distributions across international boundaries. By 2100, such shifts will occur in all countries under a ‘no mitigation’ scenario. Hotspots of transboundary shifts occur in countries that are highly dependent on fisheries for income and livelihood. The imminent onset of transboundary shifts highlight the urgency to re-formulate international fisheries policies to avoid disputes over fish stock.  Bi- and multi-lateral actions discussions should start now as lead-time for policy changes is long.

**One-Sentence Summary**

Rapid climate-induced shifts in transboundary fish stocks

**Main Text.**

In the past century, human activities have altered the physical and biogeochemical conditions of the ocean, resulting in warmer, more acidic and less oxygenated waters [@IPCC:2019tn]. Marine species' distributions reflect species' preferences for discrete environmental conditions [@Hutchinson:1957nt].As a result of these past changes, many marine species have shifted their distributions towards higher latitudes, deeper water or to follow local temperature gradients to remain within their optimal environmental "niche" [@Poloczanska:2016kk]. The biogeography of marine species is projected to continue to shift as ocean conditions change through the 21st century [@Cheung:2010dt], impacting fisheries production and economies [@Sumaila:2019ew], and compromising our capacity to reach international sustainability goals within the 2030 Agenda such as Sustainable Development Goal 14 - life below water [@UnitedNations2018; @Singh:2017ds; @Pecl:2017hu]. The projected risks of impacts can be reduced by improving the effectiveness of current fisheries management, including for species that cross international borders, which are known as ‘shared stocks’ [@Pinsky:2018cb].

The concept of shared stocks was developed following the ratification of the United Nations Convention on the Law of the Sea (UNCLOS) and the claiming of Exclusive Economic Zones (EEZs) by Coastal States [@UnitedNations:1986tl]. As defined by the United Nations' Food and Agriculture Organization, shared stocks can be classified into four non-exclusive categories: (*i*) transboundary stocks, which cross neighboring EEZs; (*ii*) straddling stocks that, in addition to neighboring EEZs, also visit the adjacent high seas; (*iii*) highly migratory stocks, mainly tunas and bill-fishes, that migrate across vast oceanic regions including both the high seas and EEZs; and (*iv*) discrete stocks that are only present in the high seas [@Munro:2004th]. This study focuses on transboundary stocks exploited by fisheries operating within EEZs. Countries are responsible for the management of stocks within their EEZs, under UNCLOS, and are encouraged to cooperate when stocks are shared [@UnitedNations:1986tl]. A recent study estimates that, globally, there are 633 transboundary fish species representing 67% of identified exploited taxa. These species yielded an annual average of 48 million tonnes of catch and USD 78 billion in fishing revenue between 2005 and 2010 [@PalaciosAbrantes:2020tv].

The effectiveness of fisheries management for transboundary species, and thus reaching the 2030 Agenda [@Pecl:2017hu], is challenged by shifts in stocks' distribution under climate change [@Pinsky:2018cb; @Pinsky:2014by; @Pinsky:2020jx; @Oremus:2020id]. In most cases, catch or fishing effort quotas for transboundary stocks are based on historical records and do not necessarily consider the full distribution range of the stock [@FredstonHermann:2018kp], nor the effects of climate change on fish stocks and associated fisheries [@PalaciosAbrantes:2020tv; @Sumby:2021gw]. While many transboundary stocks experience natural seasonal spatial variation (e.g., migration, reproductive cycle), misalignment between fisheries resources allocation and distributional shifts beyond such natural variation have previously resulted in unsustainable fishing levels and international disputes [@Miller:2013iv; @Spijkers:2017ij; @Crespo:2020cs]. Such patterns are expected to be exacerbated by continuing climate change [@Pinsky:2018cb; @Sumaila:2011bq]. The importance of transboundary stocks for both catch and fisheries revenue [@PalaciosAbrantes:2020tv] highlights their critical role in achieving the  Sustainable Development Goals (SDG) under the United Nations 2030 Agenda. Understanding when climate change will affect the sharing dynamics of transboundary stocks and the intensity of the resulting impacts is important for developing climate resilient international ocean governance and achieving the 2030 Agenda [@Pinsky:2018cb; @Link:2010ei; @Sumaila:2020tp].

Here, we employ a mechanistic population dynamic model driven by output from a comprehensive Earth System Model with ten ensemble members to project transboundary stocks’ distribution across the 280 EEZs of 198 coastal countries under a high greenhouse gas emissions scenario (RCP8.5; Methods). We then estimate the year in which the 663 transboundary species (9,132 stocks) are projected to shift their shared distribution beyond historical variation and the intensity of this shift (i.e., time of emergence; see Methods). We treat each species in an EEZ as a single stock due to the lack of more spatially resolved data to delineate the boundary of a population (15) and only consider shared stocks between neighboring EEZs (i.e., excluding the high seas).

**Identifying range shifts in the shared distribution of transboundary stocks**

Our results suggest that 4,119 transboundary stocks will experience a range shift beyond historical variability as early as 2006 and up to 2100 (Fig. \@ref(fig:toeFigure1)A). This corresponds to 68% of the studied stocks (Fig. \@ref(fig:toeFigure1)C).The average year of shift in the shared distribution of these stocks in all EEZs analyzed is projected to be 2036 (s.d. of all species, $\pm$ 28 years). Furthermore, 96% of the world’s EEZs will experience at least one transboundary stock shift between 2006 and 2100 with around 80% presenting a distributional shift across stocks by the time the UN 2030 Agenda is intended to be achieved (Fig. \@ref(fig:toeFigure1)C).

```{r toeFigure1, eval = T,  echo = F, fig.width = 9, fig.height=4, fig.pos= "H", fig.cap =  "Timeline of shifts in transboundary shared stock distribution. Dashed lines represent the year by which countries have committed to reach the full implementation of the 2030 Agenda, as well as a mid and late change in range shift, respectively."}

@@

The median year in which transboundary stocks experience a range shift varies significantly according to the geographic region of the neighboring EEZs (Kruskal-Wallis, $X^2$ = 242.11, DF = 93, *p* < 0.001; Figure \@ref(fig:toeFigure2)A). Overall, most tropical EEZs will see earlier shifts, with the EEZs of Latin America, the Caribbean and Polynesia being significantly earlier (*p* < 0.05; see Table S\@ref(tab:toeTableS2) for test statistics) than almost all other regions (Fig. \@ref(fig:toeFigure2)B). In contrast, EEZs located in temperate regions like northern Europe and eastern Asia are projected to experience much later range shifts (Table \@ref(tab:toeTableS2)).Stocks in countries such as Brazil and New Zealand are not projected to experience range shifts beyond natural variability (e.g., the transboundary index of a stock did not change beyond historical variability) (Fig. \@ref(fig:toeFigure2)A). On average, shifting transboundary stocks represented 23% (s.d. $\pm$ 22%) of yearly fishing revenue from fisheries targeting such stocks within global EEZs between 2004 and 2010 (Fig. \@ref(fig:toeFigure2)A). However, large variation exists, with shifting stocks representing less than 1% of fishing revenue in some countries (e.g., Ireland) and over 90% in others (e.g., Marshall Islands). Moreover, in some EEZs, while few transboundary stocks are projected to shift beyond historical variability, they still represent a large proportion of revenue derived from fisheries within that EEZ (e.g., Peru).

```{r toeFigure2, eval = T,  echo = F, fig.width  = 13,fig.pos= "H", fig.height =11, fig.cap =  "Year of change in the shared distribution of 4,119 transboundary stocks. A) Land polygon shows the contribution of shifting stocks to the country’s total fishing revenue from transboundary stocks. Exclusive Economic Zone polygons represent average range shifts within them. EEZs with no distributional shifts between 2006 and 2100 are represented in aqua color. B) Year of shared distribution shifts by region according to United Nations sub-regions. Points represent the median year of range shift. Horizontal dashed line represents the year 2030. N = North, S = South, W = West and E = East. Ltn. Ame. and the Car. = Latin America and the Caribbean. Aus and New Z. = Australia and New Zealand"}

@@

<!-- Set Figure, table and equations numbering according to the chapter -->
\renewcommand{\thefigure}{S.\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{S.\arabic{table}}
\setcounter{table}{0}

```{r supplements, child = '../Scripts/Supplements.Rmd', eval = T}

>>>>

---
output:
  bookdown::pdf_book:
    toc : no
    keep_tex: no
    number_sections: yes
geometry: margin=1in
editor_options:
  chunk_output_type: inline
bibliography: ../References/ToE_Ref.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{float} # Use the 'float' package
- \floatplacement{figure}{H} # Make every figure with caption = h
---

@@

#### Project's Library
packages <- c(
  "tidyverse" ,
  "png",
  "grid",
  "knitr"#,
)

@@

## Introduction

Over the last century, human activities have altered the physical and biogeochemical conditions of the ocean, including warming, acidification and reducing oxygen content [@IPCC:2019tn]. Distributions of marine species are closely related to the environment and the species’ preferences to environmental conditions (e.g., temperature, oxygen, salinity) [@Hutchinson:1957nt; @Nelson:2016kk]). As a result of climate change, many marine species have changed their distributions towards higher latitude, deeper water or have followed local temperature gradients [@Poloczanska:2016kk]. Biogeography of marine species is projected to continue to shift as ocean conditions are changing in the $21^{st}$ century under climate change [@Cheung:2010dt], impacting fisheries production and compromising our capacity to reach international sustainability goals such as Sustainable Development Goal 14 - life below water [@UnitedNations2018; @Barange:2014kr; @Singh:2017ds]. The projected risks of impacts can be reduced through improving the effectiveness of current fisheries management [@Gaines:2018sg; @Cheung:2018dh], including the fisheries management of species that cross international borders, i.e., shared stocks [@Gaines:2018sg; @Pinsky:2018cb].

The concept of shared stocks was developed following the ratification of the United Nations Convention on the Law of the Sea (UNCLOS) and the claiming of Exclusive Economic Zones (EEZs) by States [@UnitedNations:1986tl]. As defined by the United Nations’ Food and Agriculture Organization (FAO), shared stocks can be classified into four non-exclusive categories: (i) transboundary stocks, those that cross neighboring EEZs; (ii) straddling stocks, that, in addition to neighboring EEZs, also visit the adjacent high seas; (iii) highly migratory stocks, mainly tunas and bill-fishes, that migrate across vast oceanic regions including both the high seas and EEZs; and finally (iv) discrete stocks that are only present in the high seas [@Munro:2004th]. This chapter focuses on transboundary stocks exploited by fisheries operating within EEZs. While countries are responsible for the management of stocks within their EEZs, under UNCLOS, States are encouraged to cooperate when stocks are shared [@UnitedNations:1986tl]. In the previous chapter, I estimated that there are 633 transboundary fish species globally, representing 67% or the identified fished taxa, yielding an annual average of 48.5 million tonnes of catch and USD 78 billion in fishing revenue (between 2005 and 2010).

The effectiveness of fisheries management for transboundary stocks is challenged by the shifts in marine stocks distribution under climate change [@Pinsky:2018cb; @Pinsky:2014by]. In many cases, catch or fishing effort quotas for transboundary stocks are based on historical records [@Baudron:2020dj] and do not necessarily consider the biogeography of the stocks [@FredstonHermann:2018kp], nor the effects of climate change on the fish stocks and fisheries [@Sumby:2021gw, Chapter 2]. Misalignment between fisheries resources allocation and stocks’ distributional shifts have previously resulted in unsustainable harvest and international disputes [@Miller:2013iv; @Song:2017va; @Spijkers:2017ij], patterns that are expected to be exacerbated by intensifying climate change [@Pinsky:2018cb]. With future shifts in species distributions under climate change, the challenges for the  management of fisheries targeting transboundary stocks will continue to increase during the $21^{st}$ century. It is important to constrain when climate change will affect the dynamics of transboundary  stocks and the intensity of the resulting impacts in order to prepare ocean governance to meet the resource management challenges arising from shifting transboundary stocks [@Pinsky:2018cb; @Link:2010ei, Chapter 2].

The timing and intensity of climate impacts on managing transboundary fish stocks can be quantitatively examined through the concepts of ‘time of emergence’ (ToE) and ‘threat point’ (TP). Time of emergence is defined as the moment in time when a signal (e.g., future anthropogenic trend) emerges from the background noise of natural variability (e.g., historical natural variation) [@Hawkins:2012hc] and has been applied to multiple oceanic physical and biogeochemical variables [@Rodgers:2015jq; @Keller:2014jv; @Frolicher:2016dg; @Cheung:2020ja; @Schlunegger:2020gz; @Schlunegger:2019el; @Henson:2017ev]. The premise behind the time of emergence is that we can only be confident that a significant change has been detected when the signal of anthropogenic climate change is larger than the background noise of natural climate variability [@Hawkins:2012hc]. The concept of threat point comes from game theory and is defined as the minimum payoff a player is willing to receive to cooperate in a game theoretic model [@Nash:1953vr]. Game theory has been widely used to investigate and manage transboundary stocks [@Sumaila:2013hv], including the impacts of climate change on the economics of transboundary stocks between Canada and the United States [@Sumaila:2020tp].

Here, I estimate the time of emergence and threat point of the share distribution of transboundary stocks worldwide. I employ a species distribution model driven by outputs from a comprehensive Earth System Model with ten ensemble members to project the distribution of 663 transboundary species (Chapter 2) across 280 EEZs of 198 coastal countries under a high greenhouse gas emissions scenario RCP 8.5 (see Methods). I treated each species in an EEZ as a single stock due to the lack of more spatially resolved data to delineate the boundary of a population (Chapter 2) and only considered shared stocks between neighboring EEZs (i.e., transboundary stocks that exclude the high seas). I find that the distribution of many transboundary stocks has already shifted from historical natural variability and exceeded their share threat point in most neighboring EEZs, posing a threat to their sustainability and the resilience of international treaties.

## Materials and methods

### Databases and species selection

This analysis includes 633 exploited marine transboundary species previously identified to account for 80% of the catch taken from the world’s EEZs between 2005 and 2014 (Chapter 2). Each time a species was shared by a pair of neighbouring EEZs it was considered a transboundary stock [@Teh:2015gd, Chapter 2], resulting in a total of 9,132 transboundary stocks in this analysis. Using the transboundary species Atlantic cod (*Gadus morhua*) as an example, and based on the definition of transboundary stock in this chapter, the United States and Canada share a stock of Atlantic cod, and Canada and Greenland share another stock of Atlantic cod, but the United States and Greenland do not share a stock of this transboundary species. I defined the boundaries of the world’s EEZs using the Sea Around Us spatial division (updated 1 July 2015, available from http://www.seaaroundus.org), noting that it subdivides the EEZs of 198 coastal states into 280 regions (Figure \@ref(fig:toeFigureS1)) including island territories. I determined the intersections between polygons using the R statistical software package sf [@PackagesfSimpleF:2018vp]. Each EEZ was categorized by geopolitical region according to the United Nations (https://population.un.org/wpp/DefinitionOfRegions/) and biome location [@Reygondeau:2019uh]. The habitat association of each species was determined following the classification in FishBase for fish species (http://www.fishbase.org) and SeaLifeBase for invertebrates (http://www.fishbase.org). For each stock and EEZ, I used the Sea Around Us data to estimate catch and fishing revenue from fishing activities within global EEZs. I report both average catch and revenue for the last decade (2005-2014) [@Zeller2016; @Tai2017; @Sumaila:2015uc]. The monthly average US Consumer Price Index (CPI) according to the U.S. Bureau of Labor Statistic (https://www.bls.gov/cpi/) was used to standardize all monetary values to 2019 real USD.

### Projecting species distributions under climate change

I projected the distribution of marine species using a dynamic bioclimatic envelope model (hereafter called DBEM) [@Cheung:2010dt; @Cheung:2016jd]. The DBEM represents species’ physiology, habitat suitability, depth and latitudinal ranges, and spatial population dynamics as well as preferences for sea temperature, salinity, oxygen content, sea ice extent (for polar species) and bathymetry. For pelagic species, the model uses environmental variables at the surface whereas demersal species distribution is driven by ocean bottom variables. The DBEM then estimates species abundance and maximum catch potential (a proxy of maximum sustainable yield) over a $0.5^{\circ}$ x $0.5^{\circ}$ grid (see Cheung et al. [-@Cheung:2010dt; -@Cheung:2016jd] for model details). Importantly, DBEM is able to project catches for the world’s large marine ecosystems and changes in catches by EEZs that are consistent with observational-based estimates of catch from 1950 to 2016 [@Cheung:2016jd, Cheung et al. *in revision*].

The DBEM was forced with simulated ocean conditions from a ten-ensemble member simulation of the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project the distributions of the 633 species from 1951 to 2100 [@Dunne:2012bo; @Dunne:2013db; @Rodgers:2015jq]. The GFDL-ESM2M was run under historical forcing until 2005 and followed the high greenhouse gas emissions scenario, the Representative Concentration Pathway 8.5 (RCP 8.5) over the 2006-2100 period [@Riahi:2011dk]. Because the main approach of this chapter relies on understanding the spatial and temporal variation of a stocks’ distribution, I have to understand distribution variability during both the historical and the future periods, to infer differences between time frames. I do this by following an ensemble member approach where each member constitutes a different realization of the Earth system variability condition [@Rodgers:2015jq; @Frolicher:2009ge]. Thus, for my experiment, each of the ten GFDL-ESM2M ensembles were started from infinitesimally small differences in Earth system initial conditions in year 1950 resulting in a unique atmosphere and ocean state at each point in time after about three years for surface waters and eight years for subsurface waters [@Frolicher:2020cc]. By design, variations among ensemble members are then solely due to natural variability. This approach allows us to estimate the natural internal variability (e.g., background noise) and isolate the forced anthropogenic climate signal of a stock’s distribution by averaging the secular trend over all ten ensemble members.

### Calculating an index of transboundary range shift

I developed an index to evaluate range shifts in transboundary stocks under climate change (TI). This index was based on the shifts in the distribution centroid of a transboundary stock relative to the centroid of the neighboring EEZs that share this stock (Figure \@ref(fig:toeFigureS2)). The centroid of a transboundary stock was determined by the average (µ) latitude ($lat_{ts}$) and longitude ($lon_{ts}$) across the grid cells with the highest abundance within the neighboring EEZs sharing the stock e.g., Pacific halibut (*Hippoglossus stenolepis*) that is transboundary between the United States and Canada. Therefore,

<!-- $lat_{ts} = \mu (lat_{per})$ -->

<!-- $lon_{ts} = \mu (lon_{per})$ -->

\begin{equation}
\begin{split}
\label{eq:one}
lat_{ts} = \mu (lat_{per}) \\
lon_{ts} = \mu (lon_{per})
\end{split}
\end{equation}

where $lat_{per}$ and $lon_{per}$ are the latitudes and longitudes of the grid cells holding the $per^{th}$ percentile of the projected transboundary stock abundance. To focus on areas where transboundary stocks are more abundant and fishing activities are more likely to take place, I included grid cells where the projected stock abundance within the neighbouring EEZs sharing the stock was above the top 95th percentile (e.g., per = 95%). A sensitivity analysis was ran using a subset of species (n = 34) for all EEZs under different abundance levels (per = $20^{th}$, $50^{th}$ and $90^{th}$ percentile) and examined the effects of different thresholds on the calculated index value (Figure \@ref(fig:toeFigureS3)). The centroid of each EEZ was estimated using the st package in R (Figure \@ref(fig:toeFigureS1)). For each ensemble member, neighboring EEZs and transboundary stock, I computed the distance between centroids assuming the earth is a perfect sphere and ignoring geographic barriers using the geosphere package in R;

<!-- $$D_{ens} = acos(sin(lat_{eez})*sin(lat_{ts}) + cos(lat_{eez})*cos(lat_{ts})*cos(lon_{eez}-lon_{ts}))$$ -->

\begin{equation}
\label{eq:two}
D_{ens} = acos(sin(lat_{eez})*sin(lat_{ts}) + cos(lat_{eez})*cos(lat_{ts})*cos(lon_{eez}-lon_{ts}))
\end{equation}

where $lat_{eez}$ and $lat_{ts}$ are the latitudes of the EEZ and transboundary stock centroids, respectively and $lon_{eez}$ and $lon_{ts}$ are the respective longitudes. Then, for each year I calculated the transboundary index as follows:

<!-- $$TI =(\frac{D_{A,t}}{sd(D_{A,th})} - \frac{D_{B,t}}{sd(D_{B,th)}})^2$$ -->

\begin{equation}
\label{eq:three}
TI =(\frac{D_{A,t}}{sd(D_{A,th})} - \frac{D_{B,t}}{sd(D_{B,th)}})^2
\end{equation}

where $D_A$ and $D_B$ represent the distance between the distribution centroids of a stock and the centroid of the pair of neighbouring EEZs (A and B) sharing the stock for each time step from 2006 to 2100 (*t*); and *sd* is the standard deviation of the historical (*th*, 1951 - 2005) centroid distribution for $D_A$ and $D_B$. Thus, a higher TI means that a shared stock becomes more or less abundant relative to its neighbouring EEZs that is beyond what the fisheries experienced historically. From the perspective of a State’s fisheries, higher TI indicates an increase in the challenges to effective management and sustainability of the shared stock’ fisheries. In contrast, a smaller TI suggests that the sharing of the stocks between neighbouring EEZs is becoming relatively more stable; a condition that favors effective management and sustainability of the shared stocks’ fisheries. The index was smoothed to a 10-year average to reduce interannual variability.

#### Calculating the time of emergence of the transboundary index

Knowing the point in time (e.g., year) at which the distribution of a shared stock will diverge from its natural internal variability is important for informing the lead-time to which climate adaptation needs to be implemented [@Link:2010ei]. Here, I defined time of emergence as the time when the stock shared between neighbouring EEZs exceeded the historical natural internal variability as follows:

<!-- $EmergenceTIndex = \frac{\mu_{TI[t]}}{SD_{TI_{th}}}$ -->
<!-- $ToE = t when Emergence index[t] > 1$ -->

\begin{equation}
\begin{split}
\label{eq:four}
EmergenceTIndex = \frac{\mu_{TI[t]}}{SD_{TI_{th}}} \\
ToE = ToE_{t} > 1 s.d.
\end{split}
\end{equation}

where $\mu_{TI}$ is the yearly mean TI ($\mu$) across all ten ensemble members in year *t* and $SD_{TI}$ is the standard deviation of the smoothed (10-year moving average) TI across the ensemble members for the historical (*th*) reference period (1951 - 2005). I set a threshold of TI above which I considered a shared stock to have emerged from the historical internal variability of the TI. The threshold was set as one standard deviation of the variability ($SD_{TI}$) (i.e., Equation \@ref(eq:four)), representing a probability of 68% that the transboundary index has emerged from historical variability). I tested the sensitivity of the calculated emergence index to  a higher emergence threshold of two times the standard deviation (i.e., representing a probability of 95% that the index has emerged).

<!-- #### Computing a transboundary fisheries vulnerability index (VI) -->

<!-- I created a country-specific vulnerability index (VI) to examine the risk of impacts of species’ range shifts on transboundary fisheries management. The VI reflects both the rate of change of shared stocks’ distribution (indicated by the time-of-emergence of the transboundary index) and the importance of the shifting species to a country’s total catch and revenue from the species. The VI is calculated from: -->

<!-- $$ index' = (P_c + P_v)/ToE $$ -->

<!-- \begin{equation} -->
<!-- \label{eq:five} -->
<!-- VI' = \frac{(P_c + P_v)}{ToE} -->
<!-- \end{equation} -->

<!-- where $P_c$ and $P_v$ indicate the proportion accounted for by shared species in total catch and value generated from transboundary fisheries within a country’s EEZ, respectively. Here, the proportion of total catch and value are represented as the average of 2005 to 2014 within individual EEZs according to the Sea Around Us. The VI was then normalized to a 0 (less vulnerable) to 100 (more vulnerable) scale. -->

### Estimating the threat point of transboundary stock share

In game theory, cooperation over a shared resource will more likely result in a better overall outcome for the sharing parties than those operating on a non-cooperative basis [@Munro:1979dg]. However, the benefits that each player gets from a cooperative strategy must be above a minimum payoff i.e., “threat point” [@Sumaila:2013hv]. Here, I defined the threat point as the minimum required proportion of a shared stock present within an EEZ for a country to engage in cooperative management with their sharing neighbor [Chapter 2, @Sumaila:2020tp]. Any proportion below the defined threat point of a country would result in a unilateral management of the shared stock.

First, I estimated the stock share ratio (SSR) of each transboundary stock shared by neighbouring nations between 1951 and 2100. I did this by aggregating the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells in which the stock is present across neighboring EEZs, and then calculated the proportion of the stock held within each EEZ (Chapter 2). Second, I averaged the calculated proportion of stock within an EEZ held into three time periods to reduce the effects of variability. The first time period ($SSR_{th}$) spans 1951 to 2005 and was chosen to match the historical period in the GFDL ESM2M simulations, over which the model was forced with observation-based greenhouse gas, aerosol and natural external forcing [@Dunne:2012bo; @Dunne:2013db]. It was assumed to represent the historic baseline of shared distributions for each stock. I then selected two future periods; the early $21^{st}$ century as the average of 2021 to 2040 ($SSR_{te}$), and the mid $21^{st}$ century as the average of 2041 to 2060 ($SSR_{tm}$). I chose these time periods to match the challenges of achieving fisheries-related UN-SDGs such as SDG 14.4 (end overfishing), SDG 2.4 (ensure sustainable food production systems) or SDG 1.2 (poverty reduction), to be achieved by 2030 [@Singh:2017ds]. The analysis was replicated for projected stocks distributions from each of the ten ensemble members and results were averaged across ensemble members. Third, I defined a threat point for each EEZ’s stock as $SSR_{th} \pm \sigma$, where $\sigma$ is the standard deviation of $SSR_{th}$. This way, a change of SSR beyond an EEZ’s threat point happened when the future SSR exceeded one standard deviation of the historical variations of the SSR i.e when $SSR_{te} \geqq (SSR_{th} + \sigma)$ or $SSR[t]_{e,m} \leqq (SSR_{th} - \sigma)$. Finally, I estimated the percentage change in the SSR ($\Delta{SSR}$) of each future time period ($SSR_f$) relative to the historic time period ($SSR_{th}$) for each stock whose share ratio overpassed the threat point following (See Chapter 4 - Methods),

<!-- $$ \Delta{SSR_{e,m}} = \frac{(SSR_{f}-SSR_{th})}{SSR_{th}}* 100 $$ -->

\begin{equation}
\label{eq:five}
\Delta{SSR_{e,s}} = \frac{(SSR_{f}-SSR_{th})}{SSR_{th}}* 100
\end{equation}

### Statistical analysis

The time of emergence results were tested for normality (e.g., skewness, kurtosis) and performed two non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the ToE of transboundary stocks. Specifically, I tested if the habitat association of transboundary species and the geographic location of EEZs would have any effect on the time of emergence of transboundary stocks. For both cases, my null hypothesis was that there were no significant differences in the time of emergence across habitat association nor EEZs.
<!-- Moreover, I implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, I looked at habitat associations, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). I hypothesize that all of these factors would have significant influence in the ToE of transboundary species. -->
All analyses were run using the statistical software R version 3.5.2 (2018-12-20; Eggshell Igloo) with the packages data.table [@Packagedatatable:2019uh], janitor [@PackagejanitorSim:2018ur], rfishbase [@PackagerfishbaseR:2019th], sf [@PackagesfSimpleF:2018vp], sp [@PackagespClasses:2019tn], tidiverse [@PackagetidyverseE:2017vq], tidytext [@PackagetidytextTe:2019vk], gmt [@PackagegmtInterfa:2017vh] and zoo [@PackagezooSInfr:2019tm], ggrepel [@PackageggrepelAut:2020to], zeallot [@PackagezeallotMul:2018wy], viridis [@PackageviridisDef:2018us], cowplot [@CowplotStreamlined:2019wt], moments [@PackagemomensMome:2015uh] and pgirmess [@PackagepgirmessSp:2018um]. All code is available at https://github.com/jepa/TransEmergence

## Results

```{r Load_Results, eval = F,  echo = FALSE, include = FALSE}

prop_r <- read_csv("/Volumes/HALL2000/Data/TransEmergence/Results/proportion_change_2005.csv") %>% 
  gather("time_step","change",per_early,per_mid)

### Time of emergence of transboundary stocks

The average time of emergence of the transboundary index across the emerging stocks in all EEZs analyzed is projected to be 2029 $\pm$ 26 years (Figure \@ref(fig:toeFigure1)). This means that on average, the distribution of transboundary stocks will diverge from their natural internal variability by 2029. In total, the projected transboundary index of 5,745 stocks (63% of the studied stocks) will emerge from their historical variability between 2006 and 2100. Fish stocks with transboundary index that first emerged from their historical variability in each EEZ is projected to be as early as 2006 and the last one in 2100. About 55% (n = 3,154) of those emerging stocks are projected to have their transboundary index emerged between 2006 and 2020 (e.g., Peruvean anchoveta - *Engraulis ringens* - shared by Chile and Peru emerged in 2010). The number of EEZs with emerging stocks is projected to increase steeply from 87% in 2006 to 96% in 2020. After the mid-2020s, the rate of increase in the number EEZs slows from a 1% to an average of 0.7% per year. By 2081, all of the EEZs analyzed here will have at least one emerging stock (Figure \@ref(fig:toeFigure1)). Moreover, 96% of the world’s EEZs saw at least one stock having their distribution shifts between 2006 and 2020 and around one fifth (19%) of the EEZs have an average time of emergence of the transboundary index across stocks prior to 2020. Shifts in stocks’ distributions will continue to exceed historical limits steadily towards the end of the $21^{st}$ century (Figure \@ref(fig:toeFigure1)). I based the transboundary index on a transboundary stock’s centroid defined as the average grid cells that present the higher stock’s abundance (see Methods - Calculating an index of transboundary range shift). Results from the sensitivity analysis suggest that this method of estimating the time of emergence of the transboundary index is robust to the abundance level (e.g., the top 95%) of the stock (Figure \@ref(fig:toeFigureS3)).

```{r toeFigure1, eval = T,  echo = F, fig.width = 9, fig.height=4, fig.pos= "H", fig.cap =  "Time of emergence of the transboundary index by Exclusive Economic Zone (EEZ) and transboundary stock"}

@@

The median time of emergence of the transboundary index of stocks varies significantly according to the geographic region of the neighboring EEZs (Kruskal-Wallis, $X^2$ = 287.23, DF = 93, p < 0.001; Figure \@ref(fig:toeFigure2)A). Overall, most tropical EEZs will see an earlier time of emergence with the EEZs of Latin America and the Caribbean and Polynesia having significantly earlier times of emergence (p < 0.05; see Table \@ref(tab:toeTableS1) for test statistics) than almost all other regions e.g., Ecuador (ToE = 2013) and Papa New Guinea (ToE = 2018) (Figure \@ref(fig:toeFigure2)B). In contrast, EEZs located in temperate regions like northern Europe and eastern Asia have significantly later times of emergence (p < 0.05; see Table \@ref(tab:toeTableS1) for test statistics) than the rest of the world, e.g., Ireland (ToE = 2067) and North Korea (ToE = 2058). However, notable exceptions to these broad patterns exist, with  tropical EEZs displaying a later average time of emergence across stocks including the  Pacific EEZ of Honduras (ToE = 2071) and Panama (ToE = 2070) and despite being outside the tropics, Belgium (ToE = 2015) and Norway (ToE = 2019) display a relatively early average time of emergence (Figure \@ref(fig:toeFigure2)A). There are only a few cases where the transboundary index of a stock does not emerge between 2006 and 2100. In the Arctic EEZs of Canada and the US, the present chapter included three stocks: capelin (*Mallotus villosus*), saffron cod (*Eleginus gracilis*), and Pacific herring (*Clupea pallasii*). In the case of Brazil, I included four transboundary stocks, all shared with Uruguay (Chapter 2): Argentine menhaden (*Brevoortia pectinata*), Argentine shortfin squid (*Illex argentinus*), Argentine shrimp (*Pleoticus muelleri*), and pink shrimp (*Penaeus paulensis*). However, on all cases, I find that the transboundary index of these stocks did not emerge before 2100 relative to the historic (1951-2005) distribution (Figure \@ref(fig:toeFigure2)A).

```{r toeFigure2, eval = T,  echo = F, fig.width  = 13,fig.pos= "H", fig.height =11, fig.cap =  "Average time of emergence of stocks’ transboundary index. A) Land polygon shows the participation of the emerging stocks in the country’s total fishing revenue from transboundary stocks. Exclusive Economic Zone polygon represents the average time of emergence of the transboundary index across all stocks. Aqua color represents EEZs with no time of emergence between 2006 and 2100. B) Time of emergence of the transboundary index according to the United Nations sub-regions. Whiskers represent 1.5* interquartile range. Box represents interquartile range as distance between first and third quartiles. Line represents median, and black points represent outliers with values exceeding of 1.5 times the interquartile range.", fig.scap="Average time of emergence of stocks’ transboundary index.", out.extra=''}

@@

```{r FigureS1, eval = F,  echo = F, fig.width=9, fig.height=4, fig.cap =  "World Exclusive Economic Zones used in this chapter as devided by the Sea Around Us and their centroids (points)."}

img <- readPNG("../Figures/Supp1.png")
grid.raster(img)
